A significance test for forward stepwise model selection

### 2013-10-06 paper/paper.tex
Paper: expanded forward stepwise description

\usepackage{graphics,amssymb,color}
\usepackage{graphicx,subfigure}

@@

% "Title of the paper"
\title{Significance testing with forward stepwise and grouped forward
  stepwise model selection}

@@

\begin{abstract}
% perhaps de-emphasize group lasso and emphasize forward stepwise?
  We extend the work of \cite{significance:lasso} on significance
  testing for the lasso to the group lasso.  We derive the corresponding
  test statistic and its exact distribution, and show that these can be
  used with (non-penalized) forward stepwise regression and model
  selection.  For group lasso the forward stepwise procedure adds a
  group of variables in each step.  These procedures are shown to
  perform well in simulations and in a real data example.

\end{abstract}

@@

\section{Introduction}\label{sec:intro}

Forward stepwise regression is a stochastic model selection procedure
that begins with an empty model and adds the best predictor variable
in each step.  Classical significance tests fail when a model has been
selected this way and tend to be anti-conservative.  Recently,
\cite{significance:lasso} found a novel test statistic with an
appropriate null distribution that behaves well when a model has been
selected using the lasso \citep{tibshirani:lasso}.
\cite{tests:adaptive} modified and extended those results to the
group lasso \citep{grouplasso} and other adaptive regression
problems.  The present work explores the behavior of those test
statistics for models selected by forward stepwise procedures and
works out some of the detailsinvolved in applying these methods to
models with grouped variables. We find that our test statistics can be
used for valid significance testing, or for tempering the aggressive
nature of forward stepwise model selection.  The resulting methods may
be less computationally intensive than choosing a regularization
parameter for the (group) lasso by cross-validation.

\todo Write about more differences/advantages of this paper- maybe not
in this section, but: using non-asymptotic p-values, forward step is
simple(r than lasso), more?

To fix notation for the group lasso problem we consider an outcome $y
\in \real^n$.  Let an integer $G \geq 1$ be the number of groups (or
factors).  Lowercase $g$ will be used as an index for groups.  For
each $1 \leq g \leq G$ the design matrix encoding the $g$th group is
the $n \times p_g$ matrix denoted $X_g$.  Throughout most of this
paper we assume orthogonality within groups, that is $X_g^T X_g =
I_{p_g \times p_g}$ for all $g$. ( \todo Take this out if we don't
actually need it).  Let $p = \sum_{g=1}^G p_g$ be the total number of
individual variables, and let $X$ be the matrix constructed by
column-binding the $X_g$, that is  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}
X = \begin{pmatrix} X_1 & X_2 & \cdots & X_G  \end{pmatrix}
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
With each group we associate the $p_g \times 1$ coefficient vector
$\beta_g$, and write $\beta$ for the $p \times 1$ vector constructed
by stacking all of the $\beta_g$ in order.  Finally, our model for the
data is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gmodel}
y & = X \beta + \sigma \epsilon \\
   & = \sum_{g=1}^G X_g \beta_g + \sigma \epsilon
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\epsilon$ is noise.  Unless otherwise specified we will assume
i.i.d. Gaussian noise $\epsilon \sim N(0, I_{n \times n})$ throughout.
So far we have only introduced some new notation to the usual linear
regression setup, but now we define the \emph{group lasso estimator}
as the following solution to a convex problem 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gsoln}
\displaystyle \hat \beta_\lambda = \argmin_{\beta \in \real^p} \frac{1}{2} \| y - X \beta \|_2^2 +
   \lambda \sum_{g=1}^G w_g \| \beta_g \|_2
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The parameter $\lambda \geq 0$ enforces sparsity in groups: for large
$\lambda$ most of the $\beta_g$ will be zero vectors. The weights
$w_g$ are usually taken to be $\sqrt p_g$ to normalize the penalty
across groups.  Note that this includes the usual lasso estimator as a
special case when all of the groups are of size 1, since then the
penalty term is the $L^1$-norm of $\beta$.

(\todo Change this paragraph as the sections are completed)
In Section 2 we briefly review the work of (ref LTTT) concerning
significance testing for the lasso, and discuss how this result is
extended to group lasso. Section 2 also introduces more of the
terminology we use to describe the present problem. Section 3
describes a framework for forward stepwise procedures using the test
statistic derived in Section 2. In this framework groups of variables
are added in each step until some stopping criterion involving the
test statistic is satisfied. In Section 4 we show the results of some
simulations for several forward stepwise procedures, and in Section 5
we apply our method to a real data set (\todo brief description of
data).

@@

@@

\section{Forward stepwise procedures}

\begin{itemize}
  \item Brief background, cite some original papers
  \item Describe exactly how we do each step (projection, residuals, etc) 
  \item Stopping rules, cite new paper on lasso, perhaps we have
    a computation advantage
\end{itemize}

\section{Simulations}

@@

\begin{figure}
\begin{center}
\subfigure[Gaussian design]{
\label{fig:lassopvala}
\includegraphics[width=0.5\textwidth]{../figs/small_sim_gaussian_design.pdf}}
\subfigure[Large Gaussian design]{
\label{fig:lassopvala}
\includegraphics[width=0.5\textwidth]{../figs/large_sim_gaussian_design.pdf}}
%\hspace{-15pt}
\subfigure[Categorical design]{
\label{fig:lassopvalb}
\includegraphics[width=0.5\textwidth]{../figs/small_sim_categorical_design.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\end{center}
\end{figure}

@@

\input{../selection_sim_results.tex}

@@

\section{Discussion}

>>>>

\usepackage{graphics,amssymb,color}
\usepackage{graphicx,subfigure}
\usepackage[boxed]{algorithm2e}

@@

% "Title of the paper"
\title{A significance test for forward stepwise model selection}

@@

\begin{abstract}
% perhaps de-emphasize group lasso and emphasize forward stepwise?
  We apply the methods of \cite{tests:adaptive} and
  \cite{significance:lasso} on significance tests for penalized
  regression to forward stepwise model selection. For the $k$th
  variable to enter the model, previous work relied on an asymptotic
  null distribution that breaks down when grouped variables are
  entering the model and is difficult to derive. We iteratively apply
  an exact null distribution for the first (potentially grouped)
  variable on the residual from previous steps. The resulting method
  has the strengths of stepwise selection, for example parallel
  computation, but also remedies the problem of inflated test
  statistics and over-fitting.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}

Forward stepwise regression is a stochastic model selection procedure
that begins with an empty model and adds the best predictor variable
in each step.  Classical significance tests fail when a model has been
selected this way and tend to be anti-conservative.  Recently,
\cite{significance:lasso} found a novel test statistic with an
appropriate null distribution that behaves well when a model has been
selected using the lasso \citep{tibshirani:lasso}.
\cite{tests:adaptive} modified and extended those results to the
group lasso \citep{grouplasso} and other adaptive regression
problems.  The present work explores the behavior of those test
statistics for models selected by forward stepwise procedures and
works out some of the details involved in applying these methods to
models with grouped variables.  Our test statistic can be
used for valid significance testing when computed from the same
data as the model selection.  The resulting method can
be more statistically efficient than validation on held-out data, and
also more computationally efficient than penalized methods with
regularization parameters chosen by cross-validation.

(\todo Change this paragraph as the sections are completed)
In Section \ref{sec:stepwise} we establish notation and describe our
forward stepwise procedure. Section \ref{sec:testing} briefly reviews
the parts of \cite{significance:lasso} and \cite{tests:adaptive}
relevant to our significance test, and describes the group lasso which
we require for applying the test with grouped
variables. Simulation results in Section \ref{sec:simulations} using
various stopping rules for the forward stepwise procedure--including
some from \cite{sequential:fdr}--appear promising.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Forward stepwise model selection}
\label{sec:stepwise}

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of binary encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single regulatory pathway. An entire
group is included or excluded together from the final model. For
consistency we will use $g, h$ as indices rather than the usual $i, j$
throughout. Since single variables can be considered groups of size 1,
our general exposition includes non-grouped situations as a special case.

Label the outcome variable of $n$ i.i.d. measurements $y \in \real^n$. Let an integer $G \geq 1$ 
be the number of groups. For each $1 \leq g \leq G$ the design matrix encoding the
$g$th group is the $n \times p_g$ matrix denoted $X_g$, where $p_g$ is
the number of individual variables in group $g$. Define $p = \sum_{g=1}^G
p_g$ be the total number of individual variables, so $p = G$ in the
case where all groups have size 1. Let $X$ be the matrix constructed
by column-binding the $X_g$, that is  
%Throughout most of this paper we assume orthogonality within groups, that is
%$X_g^T X_g = I_{p_g \times p_g}$ for all $g$. ( \todo Take this out if
%we don't actually need it).  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}
X = \begin{pmatrix} X_1 & X_2 & \cdots & X_G  \end{pmatrix}
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
With each group we associate the $p_g \times 1$ coefficient vector
$\beta_g$, and write $\beta$ for the $p \times 1$ vector constructed
by stacking all of the $\beta_g$ in order.  Finally, our model for the
response is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gmodel}
y & = X \beta + \sigma \epsilon \\
   & = \sum_{g=1}^G X_g \beta_g + \sigma \epsilon
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\epsilon$ is noise. Unless otherwise specified we assume
i.i.d. Gaussian noise $\epsilon \sim N(0, I_{n \times n})$ and that
$\sigma$ is known.

Before we describe the forward stepwise procedure we require one last
ingredient. To each group of variables we assign a weight $w_g$. These
weights act like penalties or costs, so increasing $w_g$ makes it
\textit{more difficult} for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily, but we will only use one
particular choice, based on $p_g$, that we discuss later. With this we
are ready to describe the forward stepwise procedure Algorithm \ref{algo:fs}.

\begin{algorithm}
\DontPrintSemicolon
\KwData{An $n$ vector $y$ and $n \times p$ matrix $X$ of $G$ grouped variables}
\KwResult{Active set $A$ of variable groups included in the model}
\SetKwFunction{lsfitResidual}{lsfitResidual}
\caption{Forward stepwise procedure}
\BlankLine
$A \gets \emptyset$\;
$A^c \gets \{ 1, \ldots, G \}$\;
$r_0 \gets y$\;
\For{$counter \leftarrow 1$ \KwTo $\min (n, G)$}{
  $g^* \gets \argmax_{g \in A^c} \norm{X_g^T r_{g-1}}_2 / w_g$\;
  $A \gets A \cup \{ g^* \}$\;
  $A^c \gets A^c \backslash \{ g^* \}$\;
  $r_g \gets $\lsfitResidual{$r_{g-1}, X_{g^*}$}\;
}
\KwRet{$A$}
\label{algo:fs}
\end{algorithm}

\begin{itemize}
  \item Brief background, cite some original papers
  \item Stopping rules, cite new paper on lasso?
\end{itemize}

@@

The \emph{group lasso estimator} is the following solution to a convex
problem
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gsoln}
\displaystyle \hat \beta_\lambda = \argmin_{\beta \in \real^p} \frac{1}{2} \| y - X \beta \|_2^2 +
   \lambda \sum_{g=1}^G w_g \| \beta_g \|_2
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The parameter $\lambda \geq 0$ enforces sparsity in groups: for large
$\lambda$ most of the $\beta_g$ will be zero vectors. The weights
$w_g$ are usually taken to be $\sqrt p_g$ to normalize the penalty
across groups.  Note that this includes the usual lasso estimator as a
special case when all of the groups are of size 1, since then the
penalty term is the $L^1$-norm of $\beta$.

@@

\section{Simulations}
\label{sec:simulations}

@@

\begin{figure}
\begin{center}
\subfigure[Gaussian design]{
\label{fig:smallgauss}
\includegraphics[width=0.5\textwidth]{../figs/small_sim_gaussian_design.pdf}}
\hspace{-15pt}
\subfigure[Categorical design]{
\label{fig:smallcat}
\includegraphics[width=0.5\textwidth]{../figs/small_sim_categorical_design.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfigure[Large Gaussian lasso]{
\label{fig:largelasso}
\includegraphics[width=0.5\textwidth]{../figs/large_sim_gaussian_lasso.pdf}}
\hspace{-15pt}
\subfigure[Large Gaussian group lasso]{
\label{fig:largegrplasso}
\includegraphics[width=0.5\textwidth]{../figs/large_sim_gaussian_design.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\end{center}
\end{figure}

\input{../large_sim_results.tex}

@@

\section{Discussion}
\label{sec:discuss}
